`You Draw It': Implementation of visually fitted trends with \texttt{r2d3}

### 2022-11-01 revisions/response-to-reviewers.qmd

---
title: "Response to reviewers of JDS2207-018 "
subtitle: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
date: "2022"
# output: pdf_document
format: 
  pdf:
    include-in-header:
      - revision-edits.tex
---

## Reviewer(s)' Comments to Author

All responses to reviewers' comments are written in \response{blue}.

<!-- \todo{Need to address yet.} \dubcheck{Double check / review.} -->

### Review: Associate Editor

This is a nicely written paper about an implementation of a tool to assess visually fitted trends. This tool was inspired by a New York Times feature "You Draw It" and was developed using the r2d3 R package. The authors explain the motivation for developing the tool, provide computing details, and present results from an experiment studying how people fit trends by visualizing them on a scatter plot.

\todo{The primary element missing from the paper is how this tool and insights from the experiment might be useful for the reader. One context that came to mind was classroom use, where instructors could use the tool in lessons about multivariable relationships and regression. It would be beneficial for the authors to include more detail about how this might be used in practice.}

Some minor edits:

+ Figure 1: The gray is hard to see, particularly for the fitted line. It would also be helpful to add more details to the to the labels on the plot or to the caption to clarify what the plot is showing. Right now it is unclear unless without reading the paragraph description in the body of the text.
  + \response{Figure 1's readability was increased by changing the text on the graphic to black, darkening the gray and alpha blending of the shaded rectangle regions, making the points open circles rather than closed, and darking the gray and increasing the alpha blending of the fitted line.}
  + \response{We elaborated in the caption for figure 1 to clarify what the plot is showing: "Demonstration of a viewer's engagement with a scatterplot showing the weekly average gas prices in the United States, from 2019-June 2022. The veiwer might perform several cognitive operations such as assessing points to determine outliers, fitting a rough mental smooth/trend line (overlaying solid gray line), and using additional contextual information from long term memory to explain the variation (shaded rectangular gray regions annotated in black text)."}

+ pg. 11 - Please provide a brief description of Prolific for readers unfamiliar with company.
  + \response{A breif description of prolific, "a crowd sourcing platform for researchers to recruit participants for their online studies," was added to the Validation Study section.}

### Review: Guest Editor for the SDSS Special Issue of the JDS

I agree with the AE’s decision of “Requires Major Revision”.

+ \todo{In addition to the AE’s comments, another possible application is using the tool for staff training. There is a growing demand to train up staff given the current job market.}

+ Another comment is that I recommend the authors update figure 2 too (AE commented about figure 1), because the gray text is hard to see. 
  + \response{Figure 2 is a screenshot directly from the New York Times website used to demonstrate the use of the 'You Draw It' feature for interactive reading. While we could try to recreate this image or modify the colors, this changes the actual figure shown to readers in the New York Times.}

+ Also, figures 3 and 6 are difficult to read given there are three subplots. If there is space, perhaps the authors can reconfigure the plots, such as creating two rows (i.e., one plot in one row, and two plots in one row). 
  + \response{Figure 3 was reconstructed to provide the three subplots in vertical order spanning an entire page for better visability. As these subplots have a natural flow, we did not want to interupt this by creating a 2 + 1 subplot figure structure.}
  + \response{Figure 6 was reconstructed to provide a 1 + 2 subplot figure structure, keeping the two predicted plots from different models next to each other in the second row.}

+ Further, please note that all figures should be generated as vector images to ensure they’re high quality for final journal publication.
  + \response{Figure 1, Figure 5, and Figure 6 were generated within R and saved directly to pdf files from within the R environment.}
  + \response{Figure 2 and Figure 3 were screenshots taken from websites. We zoomed in on a high quality screen and took a screenshot; saved as high resolution png files.}
  + \response{Figure 4 was sketched using an online tool; saving this as an svg changes the appearance of the font type and we would like to keep the sketched font. We saved this at the highest resoultion the tool allowed as a png file.}

### 2022-11-01 original-2022-sdss-you-draw-it-manuscript.qmd
working on revisions

---
title: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
bibliography: references.bib
abstract: "How do statistical regression results compare to intuitive, visually fitted results? Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."
keywords: 'graphics, user interaction, regression'
format: 
  pdf:
    include-in-header:
      - edits.tex
      # - refs.tex
    keep-tex: true
    # template: template.tex
documentclass: jds
---

@@

We set out to implement ‘You Draw It’, adapted from the NYT feature, as a way to experimentally assess the patterns we see in data. 
Here, we provide technical details of the software development, utilizing interactive graphics in R [@r-software]. 
We then share results from our study which validates 'You Draw It' as a method for graphical testing on linear trends and apply an appropriate data analysis method to the participant data. We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.

@@

<!-- \svp{Describe the experiment formally. Don't be afraid to quote Mosteller's paper if you think it's necessary. Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and nonlinear analysis. Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals. Right now you have a very short summary, but you need to actually explain the full process behind it, with all of the details you'd expect from a full experimental writeup.} -->

@@

## Extension to Nonlinear Data

The analysis method presented above extends nicely beyond the linear regressions tested in @mosteller1981eye. Here we briefly demonstrate this with an example from a study examining participants ability to make forecasts for exponentially increasing data on log and linear scales. Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot]. The combination of the GAMM analysis and visual display demonstrates the strength of the 'You Draw It' method for testing statistical graphics.

@@

+ **'You Draw It' Demonstration Applet:** The shiny app used to demonstrate the 'You Draw It' method can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+ **'You Draw It' Script:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
+ **Study Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
+ **Participant Data (Linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv). 
+ **Participant Data (Nonlinear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv). 
+ **Data Analysis Code:** The code used to replicate the nonlinear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll](https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html).

>>>>

---
title: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
bibliography: references.bib
abstract: "How do statistical regression results compare to intuitive, visually fitted results? Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data. In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications."
keywords: 'graphics, user interaction, regression'
format: 
  pdf:
    include-in-header:
      - edits.tex
      - authors.tex
      # - refs.tex
    keep-tex: true
    # template: template.tex
documentclass: jds
---

@@

We set out to implement ‘You Draw It’, adapted from the NYT feature, as a way to experimentally assess the patterns we see in data. 
Here, we provide technical details of the software development, utilizing interactive graphics in R [@r-software]. 
We then share results from our study which validates 'You Draw It' as a method for graphical testing on linear trends and apply an appropriate data analysis method to the participant data. We also briefly demonstrate the use of the 'You Draw It' method and analysis on non-linear data.

@@

<!-- \svp{Describe the experiment formally. Don't be afraid to quote Mosteller's paper if you think it's necessary. Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and non-linear analysis. Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals. Right now you have a very short summary, but you need to actually explain the full process behind it, with all of the details you'd expect from a full experimental writeup.} -->

@@

## Extension to Non-linear Data

The analysis method presented above extends nicely beyond the linear regressions tested in @mosteller1981eye. Here we briefly demonstrate this with an example from a study examining participants ability to make forecasts for exponentially increasing data on log and linear scales. Along with analyzing the feedback data with the GAMM method for flexibility due to non-linear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the non-linear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot]. The combination of the GAMM analysis and visual display demonstrates the strength of the 'You Draw It' method for testing statistical graphics.

@@

+ **'You Draw It' Demonstration Applet:** The shiny app used to demonstrate the 'You Draw It' method can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
+ **'You Draw It' Script:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
+ **Study Applet:** The RShiny applet used to conduct the linear and non-linear studies can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
+ **Study Applet Code:** The code used to create the RShiny applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
+ **Participant Data (Linear):** De-identified participant data collected in the linear study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv). 
+ **Participant Data (Non-linear):** De-identified participant data collected in the non-linear study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv). 
+ **Data Analysis Code:** The code used to replicate the linear and non-linear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll](https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html).

### 2022-11-01 original-2022-sdss-you-draw-it-manuscript.tex
working on revisions

\author[1]{Emily A. Robinson\thanks{Corresponding author. Email: erobin17@calpoly.edu}}
\author[2]{Reka Howard\footnote{Email: rekahoward@unl.edu}}
\author[2]{Susan VanderPlas\footnote{Email: susan.vanderplas@unl.edu}}
\affil[1]{Department of Statistics, CalPoly-San Luis Obispo}
\affil[2]{Department of Statistics, University of Nebraska-Lincoln}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
  pdfkeywords={graphics, user interaction, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data. In this
paper, we describe the adaptation of an old tool for graphical testing
and evaluation, eye-fitting, for use in modern web-applications suitable
for testing statistical graphics. We present an empirical evaluation of
this testing method for linear regression, and briefly discuss an
extension of this method to non-linear applications.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, breakable, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

>>>>

\usepackage[dvipsnames]{xcolor} % colors
\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
\author[1]{Emily A. Robinson\thanks{Corresponding author. Email: erobin17@calpoly.edu}}
\author[2]{Reka Howard\footnote{Email: rekahoward@unl.edu}}
\author[2]{Susan VanderPlas\footnote{Email: susan.vanderplas@unl.edu}}
\affil[1]{Department of Statistics, CalPoly-San Luis Obispo}
\affil[2]{Department of Statistics, University of Nebraska-Lincoln}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
  pdfkeywords={graphics, user interaction, regression},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data. In this
paper, we describe the adaptation of an old tool for graphical testing
and evaluation, eye-fitting, for use in modern web-applications suitable
for testing statistical graphics. We present an empirical evaluation of
this testing method for linear regression, and briefly discuss an
extension of this method to non-linear applications.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
